---
title: "BINF6250 Project 03"
author: "Jason and Tiange"
date: "2025-09-26"
output: 
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
library('reticulate')
```

# Preparation

## Install packages

```{r, install_seqlogo}

reticulate::conda_install(packages=c("seqlogo"), forge=TRUE, channel="bioconda",conda="auto", additional_install_args="--solver=libmamba")
```

```{r}
reticulate::conda_install(
  packages = 'pdf2svg',
  channels = c('conda-forge', 'bioconda', 'defaults'),
  conda = 'auto',
  additional_install_args = '--solver=libmamba'
)
```

## Import modules

```{python}
import numpy as np
import seqlogo

#import function for building sequence motif & idenfitying seqs matching to motif
from data_readers import get_fasta, get_gff
from seq_ops import COMP_TRANS, reverse_complement, get_seq
from motif_ops import build_pfm, build_pwm, score_kmer, pfm_ic
```

## Set working directory

```{python}

import os
os.chdir("/courses/BINF6250.202610/students/feng.tiang/project_03")
```

# Functions

## Generate a list of sequences for test

```{python}
# Here we test your Gibbs sampler.
# You do not need to edit this or the section below. This is the Driver program

#read promoters, store in a list of strings
dir_wk = "/courses/BINF6250.202610/students/feng.tiang/project_03"
folder_seq = "data/GCF_000009045.1_ASM904v1_genomic.fna"
seq_file = f"{dir_wk}/{folder_seq}"
folder_gff = "data/GCF_000009045.1_ASM904v1_genomic.gff"
gff_file= f"{dir_wk}/{folder_gff}"

seqs = []

for name, seq in get_fasta(seq_file): # For each entry in our FASTA file
    for gff_entry in get_gff(gff_file): # For each entry in our GFF file
        if gff_entry.type == 'CDS': # If this is a coding sequence
            promoter_seq = get_seq(seq, gff_entry.start, gff_entry.end, gff_entry.strand, 50) # Extract 50 bp as a promoter
    
            """
            Because the gibbs sampling assumption is broken in just using promoters,
            and because it takes very long time to randomly progress through so many
            regions, for this example we will pre-filter for sequences that all contain
            part of the shine-dalgarno motif:
            """
            if "AGGAGG" in promoter_seq:
                seqs.append(promoter_seq)
```

Use this "seqs" list to build my Gibbs Sampler

## Build Gibbs Sampler

```{python}

import random

def GibbsMotifFinder (seqs, k, seed=42):
    '''
    Function to find a pfm from a list of strings using a Gibbs sampler
    
    Args: 
        seqs (str list): a list of sequences, not necessarily in same lengths
        k (int): the length of motif to find
        seed (int, default=42): seed for np.random

    Returns:
        pfm (numpy array): dimensions are 4xlength
        
    '''
    
    np.random.seed(seed)
    length_seqs = len(seqs)
    
    motifs_init = []
    ic_hist = []
    
    stability_window = 50  
    convergence_threshold = 1e-5
    
    for entry in range(0, length_seqs):
        # randomly select the index number within that row
        nucleotide_index = np.random.randint(0, 50-k)
        # save the k-mer into motifs_init (list)
        motif_picked = seqs[entry][nucleotide_index:nucleotide_index+k]
        motifs_init.append(motif_picked)
    
    for samp_times in range(0, 100):
        # randomly select the i-th line of seq (i ← Random(N) where N is number of DNA entries)
        entry_index_picked = np.random.randint(0, length_seqs)
        current_motif = motifs_init[:entry_index_picked] + motifs_init[entry_index_picked+1:]
        pfm = build_pfm(current_motif, k)
        ic = pfm_ic(pfm)
        ic_hist.append(ic)
        pwm = build_pwm(pfm)
        
        ith_entry = seqs[entry_index_picked]
        kmer_score_list = []
        
        for kmer_index in range(0, 50-k):
            motif_str = ith_entry[kmer_index:kmer_index+k]
            motif_score = score_kmer(motif_str, pwm)
            kmer_score_list.append(motif_score)
            
            # get key(index) from the max score, update the motifs_init
            max_score = max(kmer_score_list)
            max_index = kmer_score_list.index(max_score)
            str_update = ith_entry[max_index:max_index+k]
            motifs_init[entry_index_picked] = str_update
        
        if samp_times > 50:
            window = ic_hist[-stability_window:]
            max_change = max(window) - min(window)
        
            if max_change < convergence_threshold:
                print(f"Convergence reached at iteration {j}. Max IC change: {max_change}")
                return pfm
            
            elif samp_times == 99:
                return pfm
```

## Run 

```{python}
# Run the gibbs sampler:
promoter_pfm = GibbsMotifFinder(seqs,10)
print(promoter_pfm)

# Plot the final pfm, SAVING it to a file instead of trying to display it
seqlogo.seqlogo(seqlogo.CompletePm(pfm = promoter_pfm.T), 
                file="motif_logo.png", 
                format='png')
```

# Generative AI usage

1.  Googled "yield" (python)\
    [**`yield` pauses and saves state; `return` terminates.**](https://www.google.com/search?client=safari&sca_esv=861dc178219c5a10&rls=en&q=yield+pauses+and+saves+state%3B+return+terminates.&sa=X&ved=2ahUKEwiijL7t9PuPAxXDg4kEHahPOgQQxccNegUIiwEQAw&mstk=AUtExfBHwhHoWRb-HlKNjTlQ8FFThnClkWMlyCIYfbtR6UqIVMYwpyO_7Yh-letMs4CI9GgvoM5MWi85nvdbCu1zobFviPPOl2J37uByYeez1fO407msxk0BFSjJwPLK2Ti4Yu9k9dazT-Ypi_vUsZ7adWMvYWw-Qo2Y8o1oZw3BSMm33f4&csui=3)

    A `return` statement immediately exits the function, discarding its state. `yield` allows the function to be paused and resumed, enabling it to produce multiple values over time.

    [**`yield` creates a generator object; `return` returns a value.**](https://www.google.com/search?client=safari&sca_esv=861dc178219c5a10&rls=en&q=yield+creates+a+generator+object%3B+return+returns+a+value.&sa=X&ved=2ahUKEwiijL7t9PuPAxXDg4kEHahPOgQQxccNegUIjgEQAw&mstk=AUtExfBHwhHoWRb-HlKNjTlQ8FFThnClkWMlyCIYfbtR6UqIVMYwpyO_7Yh-letMs4CI9GgvoM5MWi85nvdbCu1zobFviPPOl2J37uByYeez1fO407msxk0BFSjJwPLK2Ti4Yu9k9dazT-Ypi_vUsZ7adWMvYWw-Qo2Y8o1oZw3BSMm33f4&csui=3)

    A function with `yield` becomes a generator function, which, when called, returns an iterable generator object. A function with `return` simply returns a single value.

2.  Gemini: save a list of selected sequences to a FASTA file

3.  Gemini: debug Test function\
    NameError: name 'COMPTRANS' is not defined. Did you mean: 'COMP_TRANS'?

4.  Gemini: random functions
